draft 1

### 2016-07-14 finalwriteup/draft.rmd
finished reviewing lit review

# Acronyms

@@

Project managers across many industries have always struggled with forecasting project costs. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by an average of 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget [@Flyvbjerg2011]. These statistics describe destructive conditions that can result in job losses, business bankruptcies, and government budget blow outs.

@@

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - even similar work may end up demanding significantly different amounts of time. Even though this is known, a fixed price is often demanded by clients before a project is awarded. If an hourly rate fee structure is proposed, clients in many industries can simply turn to another consultant willing to take the financial risks involved with fixed price contracts **(...ref...)**.

A compromise between hourly rate and fixed price fee structures is to track employee hours, and stop work once the fixed consulting fee has been exhausted. Then, negotiate further fees, or variations, with the client. This can either be the consultant's internal strategy or a strategy agreed upon by the client and consultant beforehand. If this strategy is proposed to the client before the project is awarded, it is likely the client would again turn to another consultant who is willing to offer a fixed price. If this approach is the consultant's internal strategy, their financial risk may satisfactorily be reduced. However, this behaviour may cause friction between the client and the consultant and reduce the potential for repeat business. If the project followed a predictable process but the hours inflated beyond expectation (a common situation), consultants may prefer to wear losses or endure marginable profits to preserve their client relationship. They may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction. Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed prices in complex projects. In many industries, the reality of fixed price fee structures are unavoidable and businesses conform to survive.

@@

Unfortunately, limited research is available that docuemnts the methods industry uses to estimate costs or fixed price fees for complex-projects. However, two surveys are available: one from 2000 that covers construction project cost estimation, the other is from 1992 and focuses on IT project effort estimation (which can be likened to fixed price estimation for consultants). The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practices. They found that the main method was breaking the project into detailed parts and summing up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000].

A survey by Moores and Edwards of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem.

Both surveys indicated cost and effort estimation were performed via a detailed breakdown of the tasks required to complete the project. Analysis of past project data was alluded to (comparison of similar projects) but a rigorous data analysis process was not common. Therefore it is worth investigating if a predictive model based on past data could improve the fixed price setting practice.

**you are here**

@@

@Lovallo2003 tackled the psychology behind the high failure rates of executives in predicting costs of projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stems from Kahneman's work on decision-making that won him the Nobel Prize for economics in 2002. His research argues that a person's natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplays or ignores the risk of problems that are out of the team's control. It is for this reason that it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003].

@Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project are ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. Managers rarely analyse data from similar projects even though it could significantly improve cost predictions [@Flyvbjerg2011]. This tendency led to a formalised introduction of 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011]. An example reading from this chart would be "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

![This model was implemented by the British government for a rail project that is still under completion](images/lit_review/UKrail.JPG)

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the theory of optimism bias is a credible explanation for the problem of persistant underestimation of project costs. The accompanying solution of reference class forecasting presents a promising solution given enough accurate data.

The theory of reference class forecasting was first definitively published in 2003, however, the idea of predicting project costs from previous data has been the subject of research for many preceding years and continues to this day. Although research results have often been promising, industry uptake of the idea is not yet established.

@@

In the following section, an overview of several attempts to integrate reference class forecasting will be presented followed by published theories for the lack of industry uptake.

#### Existing Models

A substantial amount of research on project cost prediction has been dedicated to the fields of software development and construction projects, along with some work in the film industry. These analyses have generally used project information from 15 to 20 large projects. A single dataset may consist of projects completed in different countries with different clients and project teams. The mathematical methods used to predict project costs vary from statistical regressions to deep-learning Neural Networks [@Love2005][@Kim2004]. Most studies report their predictive models perform well and that generally Neural Networks are more accurate than Linear Regression.

A McKinsey study outlined a case in the film industry where a motion picture company used reference class forecasting of movie revenue (cases were weighted by similarity) to decide which movies should be heavily promoted. This model improved forecasts on financial return by 135% relative to comparison of a movie to a single past case [@Flyvbjerg2014].

The existing models have shown predictive success, however the data and intention from this case study differs from previous studies. Prediction of construction cost is a different problem from consulting effort estimation and uses different variables (except the variables describing the construction project since the consulting company works in the construction industry). Software effort estimation projects are estimating time, however past studies have compared projects across many companies (obtained by surveys etc) while the case study company has access to their data only. Additionally, their data contains a rich variety of variables which cannot be exctracted from surveys. The use of their own data would also improve trust in model outputs.

#### Industry Uptake of Existing Models and Current Practices

A survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. Surveys on current industry practice in cost estimation are scarce, however, another study on the IT industry by @Jorgensen2007 argued that previously developed parametric models were not comprehensive enough to be used in industry and that expert judgment was not mobilised enough in the process of developing these tools.

Although the surveys reveal reference class models have not gained traction in either the software or construction industry, it is worth outlining their current practices in cost estimation, fee structures, and the reasons behind these. In the construction industry, infrastructure and building projects follow a traditional process, where contractual norms have developed over many decades [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. The contractor (builder) first reviews the drawings and if the plans are not detailed enough, they offer a temporary budget price. Once they feel that the drawings are sufficiently detailed, they will offer a fixed price [@Badenfelt2011]. As the job progresses however, if additional work on top of the detailed drawings is required, the builder maintains the right to charge a variation [@Badenfelt2011]. A dual source of income therefore exists, the initial proposed fee for construction along with variations due to additional materials (and time required to install the materials) that were not explicitly specified on drawings. Variations fees are generally calculated after completion the work, but the initial construction fee contains risk. Material costs are forseeable but labour and time estimates can still be under calculated, even from detailed drawings. For large infrastructure projects, small percentage miscalculations can be exceedingly expensive.

The builders' procedure of calculating a fixed price from detailed drawings and charging for variations does not translate to other consultants in the construction industry involved with the design of structures, such as engineers and architects. These consultants develop the detailed plans for the builders, and must therefore offer fixed price contracts based on low-detail preliminary drawings that sum up the client's expectations. It is also much harder for a consultant to charge variations for portions of their time spent on items that were not included in the original simple plans. Time is less tangible than physical additional materials required, and often is shrugged off by consultants for the sake of client relationships. This is despite the real and common instance of budget blow outs due to overspent time [@Harris1999]. One method of creating a variation opportunity is by stipulating a percentage fee of the final building cost that can overrule the original fixed price contract. If a consultant has spent time beyond their fixed price budget, this may account for some or in the best case, all, of their overspent time.

Construction projects generally are completed more on time and on budget when compared to IT projects as the greatest cost is the project is building materials [@Badenfelt2011]. Unlike time, building materials can be accurately calculated from detailed design drawings. Again, this omits data from consultants who are part of the building industry but charge based on their time only.

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialed. Through experimentation with contracts, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practices and the flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

It is not completely clear why industry uptake of cost estimation tools using the outside view has been weak, and it may be the case that in modern industries such as the IT industry, problems associated with cost estimation can be innovatively designed out of the contract. Possible reasons that parametric reference class forecasting models developed by researchers have not been incorporated into industry practice include:

* lack of appropriate data
    * if a project is unique, such as a national infrastructure project, it many not feel appropriate to gather sufficient data points from other countries and time periods.
* lack of understanding from industry decision makers - the prediction process was not explained well in marketing of the tools
* models were not developed in collaboration with industry
* does not feel intuitive to price from an 'outside view' of a project which disregards small project details
* insufficient time allocated to pricing that makes change burdensome

In fact, a study by @Moores1992 concluded that a lack of framework to support use of the outside view model and a failure in marketing were the two most common explanations for lack of uptake of parametric cost estimating tools. This is unfortunate as almost all research in the area indicates an outside view analysis of a job substantially improves cost estimation accuracy for fixed price projects [@Moores1992].

@@

This research explored the potential for reference class forecasting to improve cost estimation in the context of an engineering consulting company in the construction industry. In this field, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects in the case study equalled up to 25-30% of the profits from profitable projects. This demonstrates substantial room for improvement and the possiblity of multitude positive flow on effects.

Cost estimation is currently performed by first carefully reviewing preliminary drawings for a project where time costs will be calculated using a manager's personal experience with similar projects. This is often cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate past project data
* clunky data availability offered by the current database software
* lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) was employed by this company to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee timesheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, information is recorded against each project as well as each client, client contact, and each invoice.

Reference class forecasting in this study differs from previous research in that statistical analysis was performed on internal past company project data that had been progressively recorded in the company's CRM. This presented an opportunity to evaluate reference class forecasting in a unique situation, where the 'reference' projects were all internal and a comparatively large number of cases (over 2,000 projects) were available.

The CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

@@

A statistical or machine learning model predicting a measure of project financial success was developed using the available project information and by trialing numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics, however, analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs. invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study will exploit the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model for cost estimation. The benefits of this study in comparison to previous cases are that thousands of past data points are available, when previous studies used on average 15-20 cases and at most 300 [@Finnie1997][@Pai2013][@Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on company data. Managers can interrogate 'similar' data points and relate to the actual projects or speak to a colleague who was involved. This has the potential to give the 'outside view' stronger influence on the final decision.

@@

The case study provided a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it was important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that a single company is being disected in a specific industry, which limits the ability to state conclusions applicable to a broad range of companies. This can be overcome with future studies, once a range of optimal methods and variables have been determined. Other constraining factors include various forms of missing data and human data entry error.

The data set from this case study includes scarce information on the overall cost of entire construction projects. For example, managers commonly price jobs based on the budget for the entire project, of which the consultant is a minor part. The manager might typically moderate his fee by say 1% of the entire project cost. The value of the entire project has typically not been recorded although there is an allotted space in the project database for this. It would be even more beneficial for managers to record the expected total project cost at the beginning of a project versus the concluding completed cost. Another piece of data not recorded was the external factors influencing the fixed price quote including competition from similar firms that clients may have played against the case study company.

In addition to the missing data mentioned above that could be entered by workers in the future, the database also excludes the detailed information about projects that contribute to final pricing. Such details include the preliminary drawings describing the project (square meterage, number of storeys), and project summaries provided by the client to enable a quote. This level of information will never be recorded in a CRM and in many ways contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by the finer details. Given this lack of detailed information, it was not reasonable for an algorithm to predict the numeric fee for a project. It appeared much more effective to predict the profitability of a project (ie return per dollar) based on available data. This profitability is a ratio between amount invoiced and business costs. The algorithm therefore would predict the ratio of these two values based on fees and costs generated using the established method of fee proposals and time expenditure in historical cases.

Finally, CRM data is entered by employees manually as time passes, jobs are invoiced, and projects are finally completed. Because the data is input by people, it is susceptible to input error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

Although this case study presented an opportunity to test the predictive power held within the organisations' existing CRM data, limitations exist which must be understood and built into the framework of the problem. These include missing data that could have been entered for each project, but historically has not been required, missing data that is very detailed and will never be available, and the potential for a small degree of human error to be embedded within the data.

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task persuading managers to use reference class forecasting models in practice and remove themselves from summaries based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models. This research examines the potential for a company's internally generated CRM data to be used in creating a reference class forecasting model. Familiarity with the data has the potential to influence managers' decisions in the company more, however its limitations must be understood in order to create a valid model.

@@

**General Aim** Use statistical techniques to model the profitability of projects for consulting businesses using their internal CRM data. Research will focus on a case study Engineering consulting company that offers their expert advice (in the currency of time) to business clients.

The project outcomes are intended to assist the business in predicting project profitability before engagement. Several statistical and machine learning techniques were tested, compared and refined.

@@

Past research has tackled cost estimation in various industries, from different angles and with a range of data, however the problem has rarely been addressed using a single company's internal data. In the literature, dozens of projects from other companys and even countries are collected and analysed. This thesis uses the internal CRM data from a company which contains over 2,000 past job cases - far more than the usual cost estimation study. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry that estimates and must stand by their fixed fees before a project is won. This differs from other studies estimating actual construction costs of projects and is likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to Neural Networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialed which has not yet been applied to the cost estimation problem.

Finally, this study takes the model one step further than previous studies and analyses concrete profit improvements for the company if the model output were to be used in certain decision-making scenarios. A few possibilities are presented as ways to take model results into account with business decisions and where possible, final profit calcuations are presented. It is anticipated that this kind of thought experiment would further push industry to seriously consider the worth in adopting statistical cost estimation models.

@@

The remainder of this thesis has the following structure: the next chapter is a thorough review of the literature relating to this thesis to date. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of effective statistical and machine learning methods.

>>>>

# List of Abbreviations

@@

Project managers across many industries have always struggled with forecasting project costs. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by an average of 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget [@Flyvbjerg2011]. Another review of 6 IT project surveys between 1984 and 1994 by @Molokken2003 revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. These statistics describe difficult conditions that can result in job losses, business bankruptcies, and government budget blow outs. A survey from @Moores1992 asked software managers whether they saw effort estimation as a problem, and 91% responded 'yes'.

@@

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - even similar projects may end up demanding significantly different amounts of time. Although this is known, a fixed price is often demanded by clients before a project is awarded. If an hourly rate fee structure is proposed, clients in many industries can simply turn to another consultant willing to take the financial risks involved with fixed prices **(...ref...)**. Two strategies to guard against fixed price budget blowouts are detailed below along with their limitations.

A compromise between the hourly rate and fixed price structure is to track employee hours, and stop work once the fixed consulting fee has been exhausted. Then, negotiate further fees, or variations, with the client. This can either be agreed upon with the client beforehand or the consultant's internal strategy. If this strategy is proposed to the client up front, it is likely the client would again turn to another consultant who is willing to offer a fixed price. If this approach is the consultant's internal strategy, their financial risk may satisfactorily be reduced. However, the behaviour may cause friction between the client and the consultant, therefore reducing the potential for repeat business. In the event of hours inflating beyond expectation, consultants may choose to wear losses or endure marginable profits to preserve their client relationship. They may reason it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction.

Another strategy for managing this risk is by stipulating a specific one-off variation opportunity in the contract. This applies to consulting work that falls under the client's larger project budget. The variation would specify a percentage fee of the overall project cost that can overrule the original fixed price. The purpose of this rule is to guard against the case where the project creeps in size, and the consultant finds themselves gradually undertaking more work and more work for the same price even though the client was happy to spend more on the overall project. This may account for some or in the best case, all, of their overspent time, although again a savvy client may turn to other consultants that agree to omit variation opportunities.

Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed prices in complex projects. In many industries, the reality of fixed price fee structures are unavoidable and businesses conform to survive.

@@

Unfortunately, limited research is available that documents the methods industry uses for complex-projects. However, two surveys are available: one from 2000 that covers construction project cost estimation, the other from 1992 and focuses on IT project effort estimation (which can be likened to fixed price estimation for consultants). The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practices. They found that the main method was breaking the project into detailed parts and summing up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000].

A survey by @Moores1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem.

Both surveys indicated cost and effort estimation were performed via a detailed breakdown of the tasks required. Analysis of past project data was alluded to (comparison of similar projects) but a rigorous data analysis process was not common. Therefore it is worth investigating if a predictive model based on past data could improve the fixed price setting practice.

@@

Although the surveys indicated industry did not analyse past project cost data, the idea of using this data (in the same way) has been studied before. Research exists that tested people's estimation accuracy with and without comparitive historic data with interesting results. This is summarised below along with the resulting data-driven models.

@Lovallo2003 delved into the psychology behind why executives so often severely underestimate costs of larger projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stemmed from Kahneman's work on decision-making that won him the Nobel Prize for economics in 2002. His research argued that a person's natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplays or ignores the risk of problems that are out of the team's control. This is why it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003].

@Lovallo2003 call the practice of analysing a project based on project details without considering unforseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011]. An example reading from the chart is as follows: "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

![This model was implemented by the British government for a rail project that is still under completion [@Flyvbjerg2011]](images/lit_review/UKrail.JPG)

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the optimism bias theory credibly helps to explain the persistant problem of project cost underestimation.

@@

Although @Flyvbjerg2011's reference class forecasting theory and the psychology behind it was first published in 2003, predicting project costs using previous data has been researched for many preceding years and continues to this day. In this section, a brief overview of other reference class forecasting models is presented (they are reviewed in greater detail in the literature review) followed by an analysis of why industry has not yet adopted them.

Most cost estimation or reference forecasting analytical models focused on software development and construction projects, along with some work on film projects. These analyses generally used information from 15 to 20 large projects and the dataset summarised projects from around the world with different clients and teams. Applied mathematical methods varied from statistical Regressions to deep-learning Neural Networks [@Love2005][@Kim2004]. Most studies reported that their predictive models performed well and that Neural Networks were generally more accurate than Linear Regression.

#### Industry Uptake of Existing Models

Despite the predictive success in statistical and machine learning cost estimation models as well as the revealing outcomes from 'inside' and 'outside' view studies, reference class forecasting there has been a lack of industry uptake. Two reports reviewing cost estimation in the software industry investigated reasons why data-driven forecasting models have not been translated to industry. The key causes were:

* A lack of a decision framework to support manager's use of the 'outside view' model [@Moores1992]
* A failure in marketing of the potential for the models to improve business outcomes [@Moores1992]
* The products of the research were not comprehensive enough to be used in industry [@Jorgensen2007]
* Expert judgment was not mobilised enough in the process of developing the tools and the resulting product felt disconnected to actual requirements [@Jorgensen2007].

Reports on other industries were not available, however the insights are general and may explain the lack of adoption in other industries (such as the construction industry).

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractural arangements. Through experimentation, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is more modern, the new methodology was able to sweep across the field and could be proven to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high level of legal responsibility and due diligence.

This project aims to assist a company in the construction industry in managing problematic fixed price projects, since fixed price contracts are unavoidable. The blocks preventing industry adoption, such as providing a clear decision framework and incorporating expert judgment will be carefully addressed.

@@

The case study company is an engineering consulting company in the construction industry. In their profession, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects equalled 25-30% of their profits from profitable projects. This demonstrates the magnitude of their cost estimation problem and the substantial room for improvement that a data-driven model may facilitate.

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate past project data
* clunky data availability offered by their database software
* lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) is currently employed to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee timesheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, technical information is recorded against each project as well and the CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

@@

The CRM software readily provides simple output statistics, however, analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs. the invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997][@Pai2013][@Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may give the presented 'outside view' a stronger meaning and more potential to influence the final fee proposal.

@@

The case study provides an opportunity to test whether passively collected CRM data can be mobilised into a predictive model to improve fixed price cost estimation. However, there are some limitations in the study including the size of the study, missing data, and errors in data entry. This section addresses how the limitations can be addressed either in future work or within the research framework.

The most obvious limitation is that the work focuses on a single company in a specific industry, which limits the application of conclusions to a broader range of companies. This can be overcome by first determining the optimum method and variables for this case and then testing them on future case study companies.

Another limitation is the type of unavailable data. Information on overall costs of construction projects was scarce. This is relevant because managers sometimes price jobs based on the budget for the entire project, of which the consultant is a minor part. For example, engineers may moderate their fee by 1% of the entire project cost. In the future, it would be beneficial for managers to record the expected total project cost at the beginning of a project versus the concluding completed cost. Another piece of data not recorded was the external factors influencing the fixed price quote including known competition from similar firms.

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. return per dollar. This measure represents the ratio of the amount invoiced to business costs, and can easily be compared between projects of different sizes.

Finally, all CRM data is entered by employees manually. Because the data is input by people, it is susceptible to some error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, cause substantial financial losses to businesses and governments. Studies have explored human nature and our tendency to optimistically assess our capabilities with respect to a given task. In the case of cost estimation, the optimism can result in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project. This was done by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve cost estimation accuracy, however industry has not adopted them. Managers tend to predict costs based on intricate details of a project in favour of reference class forecasting. This may be due to poor marketing of the cost estimation tools, a lack of a decision framework in relation to the tools, and a lack of collaboration with industry during model development.

@@

**General Aim** Use statistical techniques to model the profitability of projects for consulting businesses using their internal CRM data. Research will focus on a case study Engineering consulting company that offers their expert advice (in the currency of time) to business clients. 
The project outcomes are intended to assist the business in predicting project profitability before engagement. Several statistical and machine learning techniques were tested, compared and refined.

@@

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the dataset is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending was tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation datasets generally consisted of dozens of projects from a range of a companys and even countries. This thesis case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset, ensemble tree methods are applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, Case Based Reasoning, and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression (**ref**). Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

Finally, this study propels the predictive model one step further than other studies by analysing bottom-line profit improvements that could result from the research. A few possibile ways to integrate model results into business decision-making scenarios are presented, and where possible, final profit increases are evaluated. This kind of thought experiment is designed to present a clear case for industry when evaluating if and how to adopt cost estimation models.

@@

The next chapter, Chapter 3 is a thorough review of the literature to date that relates to this thesis. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of effective statistical and machine learning methods.

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios. Further applications were also considered.

The research outlined in the method were divided into four stages, each of which was assigned a chapter containing the results and discussion of that portion:
